---
title: "Loan Prediction EDA"
output: html_document
date: "2023-10-24"
---

```{r}
library(dplyr)
library(ggplot2)
```


```{r}

file1 <- read.csv("./Data/Loan_Default1.csv")
file2 <- read.csv("./Data/Loan_Default2.csv")

loan_df <- bind_rows(file1, file2) %>% 
  mutate(Status = as.factor(Status))
```


# Looking at numerical variables

```{r}
num_df <- loan_df %>% 
  select_if(is.numeric) 

cat_df <- loan_df %>% 
  select_if(is.factor)


# Numerical Variables
head(num_df)
```

## What does the loan default look like

```{r}
loan_df %>% ggplot(aes(x=Status)) +
  geom_bar() +
  theme_bw()

loan_df %>% 
  group_by(Status) %>% 
  summarise(n=n()) %>% 
  mutate(freq=n / sum(n))

```

We have around a 75-25 split for no default to default occurences. A successful model would have to achieve an accuracy above 75%


## Which numerical variables have missing values

```{r}
num_df %>% summarise(across(everything(), ~ sum(is.na(.))))
```

### What do the numerical variables mean?

Self Explanatory: ID, year, loan_amount, rate_of_interest, 

(?) Symbol indicates not sure of the definition

Interest_rate_spread: Profit that the bank obtains by lending you the money on the mortgage interest rate
Upfront_charges(?): Could be various charges incurred with the transaction
term: Loan term in months
property_value: The value of the property which the loan is for
income: income of the buyer
Credit_Score: Credit score of the buyter
LTV: measure comparing the amount of your mortgage with the appraised value of the property
Status: Status of loan default
dtir1(?): Possibly debt-to-income ratio?


#### Numerical variables to remove

```{r}
loan_df %>% filter(Status == 1 & !is.na(rate_of_interest)) %>% count()
loan_df %>% filter(Status == 1 & !is.na(Interest_rate_spread)) %>% count()
loan_df %>% filter(Status == 1 & !is.na(Upfront_charges)) %>% count()
```
Rate_of_interest is missing for all rows except for 200 when Status equals 1
Interest_spread_rate is missing for all rows when Status equals 1
Upfront charges is missing for all rows except 153 when Status equals 1

We likely want to drop these variables since the available information we have when the Loan defaults is very low

#### Term

```{r}
loan_df %>% 
  filter(!is.na(term)) %>% 
  ggplot(aes(x=term)) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))
```
There doesn't seem to much of a difference between the distributions of default and non-defaulted loans. It is not likely that this variable will contribute significantly to predicting defaults.

### Property value, Income and Loan Amount
```{r}
loan_df %>% 
  filter(!is.na(property_value)) %>% 
  filter(property_value < 3000000) %>% 
  ggplot(aes(x=property_value), col = Status) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))


loan_df %>% 
  filter(!is.na(property_value)) %>% 
  filter(property_value < 3000000) %>% 
  ggplot(aes(x=property_value), col = Status) +
  geom_boxplot() +
  theme_bw() + 
  facet_grid(vars(Status))
```

```{r}
loan_df %>%  
  summarise(max(property_value))
```
When we look at the max property value, a 16.5 million dollar property does not seem impossible so we will keep the outliers.

```{r}
loan_df %>% 
  filter(!is.na(income)) %>% 
  filter(income < 40000) %>% 
  ggplot(aes(x=income), col = Status) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))


loan_df %>% 
  filter(!is.na(income)) %>% 
  filter(income < 40000) %>%
  ggplot(aes(x=income), col = Status) +
  geom_boxplot() +
  theme_bw() + 
  facet_grid(vars(Status))
```


```{r}
loan_df %>%  
  summarise(max(income, na.rm=TRUE))
```
When looking at the max income, a $578,580 income does not seem impossible so we will keep the outliers.


```{r}
loan_df %>% 
  filter(!is.na(loan_amount)) %>% 
  filter(loan_amount < 100000) %>% 
  ggplot(aes(x=loan_amount), col = Status) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))


loan_df %>% 
  filter(!is.na(loan_amount)) %>% 
  filter(loan_amount < 1000000) %>%
  ggplot(aes(x=loan_amount), col = Status) +
  geom_boxplot() +
  theme_bw() + 
  facet_grid(vars(Status))
```

For all three of these variables, we do a similar analysis.

First we remove some of the outliers to have a better look at the bulk of the distribution. All three variables show a similar story where the distributions look similar for both the default/non-default populations, but there is a slight skew for larger values for the non-default population. This makes sense because higher income people would buy more expensive properties and have larger loan amounts. The hypothesis is that higher income people should also default less.


### Credit Score


```{r}
loan_df %>% 
  filter(!is.na(Credit_Score)) %>% 
  ggplot(aes(x=Credit_Score), col = Status) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))
```
Credit score surprisingly does not show much of a difference in distrubtion between the two populations. There seems to be a large spread for all values of credit score for the defaulted and non-defaulted populations.

### LTV and dtir1

```{r}
loan_df %>% 
  filter(!is.na(LTV)) %>% 
  filter(LTV < 1000) %>% 
  ggplot(aes(x=LTV), col = Status) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))

loan_df %>% 
  filter(!is.na(LTV)) %>% 
  filter(LTV < 1000) %>%
  ggplot(aes(x=LTV), col = Status) +
  geom_boxplot() +
  theme_bw() + 
  facet_grid(vars(Status))
```
The default population shows a slight skew towards higher LTV ratios which makes sense. We also see that the default population has much more higher outliers. Theoretically the loan amount shouldn't exceed 100% of the property value and it definitely shouldn't exceed over a 1000%. It's possible that this is a certain loan type which we will look at in the categorical section. 

```{r}
loan_df %>% 
  filter(LTV > 200) %>% 
  select(loan_amount, property_value, LTV)

```
Interestingly enough, it seems the LTV is calculated propertly and we have extremely large loan amounts for certain properties. Most of thse properties are lower values with very high loan amounts. For the analysis we may want to remove some of the observations with extremely high LTV ratios.


```{r}
loan_df %>% 
  filter(!is.na(dtir1)) %>% 
  ggplot(aes(x=dtir1), col = Status) +
  geom_histogram(aes(y = stat(density))) +
  theme_bw() + 
  facet_grid(vars(Status))

loan_df %>% 
  filter(!is.na(dtir1)) %>% 
  ggplot(aes(x=dtir1), col = Status) +
  geom_boxplot() +
  theme_bw() + 
  facet_grid(vars(Status))
```
dtir1 also shows a similar story to LTV. There is a slight skew for higher dtir1 values in the default population. If dtir1 is the debt-to-income ratio then this story would make sense.


## Categorical Predictors

```{r}
cat_df <- loan_df %>% 
  select(!is.numeric & !is.factor) %>% 
  mutate_all(na_if,"") %>% 
  bind_cols(loan_df %>% select(Status))

cat_df %>% 
  summarise(across(everything(), ~ sum(is.na(.))))
```
There are alot less missing values in the categorical variables than the numerical predictors.


### What do the categorical variables mean?

Self Explanatory: age, Gender, Status, Region

(?) Symbol indicates not sure of the definition

loan_limit (?): 
approv_in_adv: Approval of loan in advance
loan_type: Type of loan
loan_purpose: Purpose of the loan
Credit_Worthiness: a measure of how likely you will default on your debt obligations according to a lender's assessment, or how worthy you are to receive new credit.
open_credit: Open credit line
business_or_commercial: Business or commercial building
Neg_ammortization: the amount you owe will still go up because you are not paying enough to cover the interest.
interest_only: a type of mortgage in which the mortgagor (the borrower) is required to pay only the interest on the loan for a certain period
lump_sum_payment: Upfront payment to the loan, down payment
construction_type (?):
occupancy_type: There are three different classifications of properties when it comes to financing: Primary Residences, Second Homes, and Investment Properties.
Secured_by: Home or land as collateral
total_units: Total housing units in the property
credit_type(?): 
co.applicant_credit_type (?): 
submission_of_application (?):
Security_Type (?):

```{r}
plot_bar <- function(df,var){
  df %>% 
    filter(!is.na({{var}})) %>% 
    ggplot(aes(x={{var}}), col = Status) +
    geom_bar(aes(y = ..prop.., group = 1)) +
    theme_bw() + 
    facet_grid(vars(Status))
}
```

#### Heavily skewed to one category
```{r}
plot_bar(cat_df, construction_type)
plot_bar(cat_df, Secured_by)
plot_bar(cat_df, Security_Type)
```
These three variables have observations mostly or all in one category so they would not be very useful in the model. We can likely remove these variables.


#### Not too much variance between populations

```{r}
plot_bar(cat_df, loan_limit)
plot_bar(cat_df, approv_in_adv)
plot_bar(cat_df, Credit_Worthiness)
plot_bar(cat_df, open_credit)
plot_bar(cat_df, interest_only)
plot_bar(cat_df, occupancy_type)
plot_bar(cat_df, total_units)
```
These variables have do not show much variance in distrbition between the two populations. We can keep these in the model to check if they have significance.


#### Control variables
```{r}
plot_bar(cat_df, age)
plot_bar(cat_df, Gender)
plot_bar(cat_df, Region)
```
These variables also have slight variation but we can keep them in the model because they are control variables.


#### Slight variance
```{r}
plot_bar(cat_df, loan_type)
plot_bar(cat_df, loan_purpose)
plot_bar(cat_df, business_or_commercial)
plot_bar(cat_df, Neg_ammortization)
plot_bar(cat_df, lump_sum_payment)
```

These variables show slightly more variation between the distributions but still not too much variance.

#### Lots of variance between populations
```{r}
plot_bar(cat_df, credit_type)
plot_bar(cat_df, co.applicant_credit_type)
plot_bar(cat_df, submission_of_application)
```

These variables have lots of variance between the two population distributions.

## Summary

#### Control Variables:

age, Gender, Region

#### Variables to potentially remove:

rate_of_interest, Interest_rate_spread, Upfront_charges, term,
construction_type, Secured_by, Security_Type


#### Variables that might not be significant

Credit_Score, loan_limit, approv_in_adv, Credit_Worthiness, open_credit, interest_only, 
occupancy_type, total_units


#### Variables that might have some signficance


property_value, income, Loan Amount, LTV, dtir1
loan_type, loan_purpose, business_or_commercial, Neg_ammortization, lump_sum_payment


#### Variables that probably are significant

Property value, Income and Loan Amount
credit_type, co.applicant_credit_type, submission_of_application




# Testing which variables are significant

One way to check which variables are the most significant to the model is to run a logistic regression on different samples of the data. Then we can see how many times a certain variable was significant throughout the consecutive runs.


```{r}
loan_df2 <- loan_df %>%
  ## Removing several category variables because they only contain one value
  select(-ID, -year, -construction_type, -rate_of_interest, -Interest_rate_spread, -Upfront_charges, -term,
-construction_type, -Secured_by, -Security_Type) %>% 
  mutate(Status = factor(Status))


imputed_set <- 
  recipe(Status ~ ., data = loan_df2) %>%
  # Use bagged models to fill in missing data
  step_impute_bag(income, dtir1, LTV, property_value, loan_limit, approv_in_adv,
                  loan_purpose, Neg_ammortization, age, submission_of_application, impute_with=imp_vars(all_predictors())) %>% 
  prep() %>% 
  bake(loan_df2)

```

```{r}
significant_vars <- c()

set.seed(1234)
for(i in 1:40) {
 sample <- imputed_set %>% 
  slice_sample(prop = 0.05) 
 
 glm1 <- glm(Status ~ ., data=sample, family='binomial') 
 coeffs <- coef(summary(glm1))
 
 significant_vars <- c(significant_vars, names(coeffs[,4])[coeffs[,4] < 0.05])
}

tibble(coeffs = significant_vars) %>% 
  group_by(coeffs) %>% 
  summarise(num_significant=n()) %>% 
  arrange(-num_significant)
```

We can see the variables that appeared significant the most amount of times out of 40 different runs. The categorical variables loan_type, lump_sum_payment, occupancy_type, and submision of application all appears to be the most significant vaiables.

Income, property value, and LTV are the next most significant variables.